---
title: "Shooting Project"
author: "Jessie Black"
date: "2023-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Shooting Project
## Dataset

The Historic NYPD Shooting Incident Data lists all shooting incidences in NYC from 2006 through the end of the most recent calendar year (2022 at time of publishing). 

Every record represents a single shooting incident and includes the following: 

- location event occurred
- time event occurred
- suspect information
- victim demographics


## Libraries Used
```{r lib}
library(tidyverse)
library(lubridate)
```

## Data Collection and Inspection

```{r import}
url = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv"
shooting_data <- read_csv(url)
summary(shooting_data)
head(shooting_data)
```
### Initial observations of variables:

- INCIDENT_KEY: numeric value (unique identifier)
- OCCUR_DATE: character value in form MM/DD/YYYY -- should be converted to date.
- OCCUR_TIME: numeric in form hh:mm:ss
- BORO: character value
- PRECINCT: numeric value
- JURISDICTION_CODE: numeric value 
- LOCATION_DESC: Character value
- STATISTICAL_MURDER_FLAG: logical
- PERP_AGE_GROUP: character. NA values numerous, presumably due to unknown perpetrator.
- PERP_SEX: character -- should be a factor. NA values numerous, presumably due to unknown perpetrator.
- PERP_RACE: character. NA values numerous, presumably due to unknown perpetrator.
- VIC_AGE_GROUP: character
- VIC_SEX: character - should be a factor
- VIC_RACE: character
- X_COORD_CD: numeric
- Y_COORD_CD: numeric
- Latitude: numeric
- Longitude: numeric
- Lon_Lat: character

### Tidying Data

1. Convert VIC_RACE and VIC_SEX to factors (wait to do the same for PERP_ after addressing NA values)
2. Convert OCCUR_DATE to date format
3. Remove unnecessary columns for our analysis (x & y coordinates, latitude, longitude, precinct, jurisdiction code)

```{r tidy_data}
shooting_data <- shooting_data %>% 
  # Convert victim's sex and race columns to factors, convert date from number to date
  mutate(OCCUR_DATE=mdy(OCCUR_DATE),VIC_SEX = as.factor(VIC_SEX),VIC_RACE = as.factor(VIC_RACE)) %>%
  # Exclude the following columns from final data set
  select(-c(X_COORD_CD, Y_COORD_CD, Latitude, Longitude, Lon_Lat, PRECINCT, JURISDICTION_CODE))
```

### Missing Values
Displaying missing values for all remaining columns:
```{r find_missing_values}
col_na <- colSums(is.na(shooting_data))
data.frame(na_count = col_na)
```
#### Analysis 
It makes sense for similar (large) amounts of NA values to exist for PERP_AGE_GROUP, PERP_SEX, and PERP_RACE -- presumably, these were incidents where the perpetrator was unidentified. There are slightly more observations with a missing PERP_AGE_GROUP than for PERP_SEX and PERP_RACE (which both had the same value), presumably because it is easier for a victim to guess at an attacker's sex and race than their age range if they were not apprehended. We could change these to "unknown" so that they may qualify as a separate factor.

There are also a large amount of values missing for LOCATION_DESC, which is more confusing, as this is more likely to be identified. The simplest explanation is that it is simply not considered a vital aspect of an incident report and is more frequently left out. As over 50% of the observations have NA as LOCATION_DESC, we could shift it to "unknown" but in this case I will instead drop the column to remove the risk of skewed data.

``` {r final_tidy}
shooting_data <- shooting_data %>% 
  # Convert any NA values to "unknown" in following three columns
  mutate(PERP_SEX = as.factor(ifelse(is.na(PERP_SEX), "unknown", PERP_SEX)), 
         PERP_RACE = as.factor(ifelse(is.na(PERP_RACE), "unknown", PERP_RACE)),
         PERP_AGE_GROUP = ifelse(is.na(PERP_AGE_GROUP), "unknown", PERP_AGE_GROUP)) %>%
  # Exclude LOCATION_DESC column from final data set
  select(-LOCATION_DESC)
```

## VISUALIZATION AND ANALYSIS
TODO: add at least 2 diff visualizations and some analysis
TODO: Consider any additional questions this raises to investigate

## ADD BIAS IDENTIFICATION
TODO: Write conclusion, including any possible bias sources. Identify personal bias & how mitigating.
