---
title: "Shooting Project"
author: "Jessie Black"
date: "2023-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Shooting Project
## Dataset

The Historic NYPD Shooting Incident Data lists all shooting incidences in NYC from 2006 through the end of the most recent calendar year (2022 at time of publishing). 

Every record represents a single shooting incident and includes the following: 

- location event occurred
- time event occurred
- suspect information
- victim demographics


## Libraries Used
```{r lib}
library(tidyverse)
library(lubridate)
```

## Data Collection and Inspection

```{r import}
url = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv"
shooting_data <- read_csv(url)
summary(shooting_data)
head(shooting_data)
```
### Initial observations of variables:

- INCIDENT_KEY: numeric value (unique identifier)
- OCCUR_DATE: character value in form MM/DD/YYYY -- should be converted to date.
- OCCUR_TIME: numeric in form hh:mm:ss
- BORO: character value
- PRECINCT: numeric value
- JURISDICTION_CODE: numeric value 
- LOCATION_DESC: Character value
- STATISTICAL_MURDER_FLAG: logical
- PERP_AGE_GROUP: character. NA values numerous, presumably due to unknown perpetrator.
- PERP_SEX: character -- should be a factor. NA values numerous, presumably due to unknown perpetrator.
- PERP_RACE: character. NA values numerous, presumably due to unknown perpetrator.
- VIC_AGE_GROUP: character
- VIC_SEX: character - should be a factor
- VIC_RACE: character
- X_COORD_CD: numeric
- Y_COORD_CD: numeric
- Latitude: numeric
- Longitude: numeric
- Lon_Lat: character

### Tidying Data

1. Convert VIC_RACE and VIC_SEX to factors (wait to do the same for PERP_ after addressing NA values)
2. Convert OCCUR_DATE to date format
3. Remove unnecessary columns for our analysis (x & y coordinates, latitude, longitude, precinct, jurisdiction code)

```{r tidy_data}
shooting_data <- shooting_data %>% 
  # Convert victim's sex and race columns to factors, convert date from number to date
  mutate(OCCUR_DATE=mdy(OCCUR_DATE),VIC_SEX = as.factor(VIC_SEX),VIC_RACE = as.factor(VIC_RACE)) %>%
  # Exclude the following columns from final data set
  select(-c(X_COORD_CD, Y_COORD_CD, Latitude, Longitude, Lon_Lat, PRECINCT, JURISDICTION_CODE))
```

### Missing Values
Displaying missing values for all remaining columns:
```{r find_missing_values}
col_na <- colSums(is.na(shooting_data))
data.frame(na_count = col_na)
```
#### Analysis 
It makes sense for similar (large) amounts of NA values to exist for PERP_AGE_GROUP, PERP_SEX, and PERP_RACE -- presumably, these were incidents where the perpetrator was unidentified. There are slightly more observations with a missing PERP_AGE_GROUP than for PERP_SEX and PERP_RACE (which both had the same value), presumably because it is easier for a victim to guess at an attacker's sex and race than their age range if they were not apprehended. We could change these to "unknown" so that they may qualify as a separate factor.

There are also a large amount of values missing for LOCATION_DESC, which is more confusing, as this is more likely to be identified. The simplest explanation is that it is simply not considered a vital aspect of an incident report and is more frequently left out. As over 50% of the observations have NA as LOCATION_DESC, we could shift it to "unknown" but in this case I will instead drop the column to remove the risk of skewed data.

``` {r final_tidy}
shooting_data <- shooting_data %>% 
  # Convert any NA values to "unknown" in following three columns
  mutate(PERP_SEX = as.factor(ifelse(is.na(PERP_SEX), "unknown", PERP_SEX)), 
         PERP_RACE = as.factor(ifelse(is.na(PERP_RACE), "unknown", PERP_RACE)),
         PERP_AGE_GROUP = ifelse(is.na(PERP_AGE_GROUP), "unknown", PERP_AGE_GROUP)) %>%
  # Exclude LOCATION_DESC column from final data set
  select(-LOCATION_DESC)
```

## VISUALIZATION AND ANALYSIS
### Understanding the Data
To start, I want to see the relationship between location and the total number of incidents. I will use BORO and COUNT(INCIDENT_KEY) for this.
``` {r incidents_by_boro}
# Create new data frame with just the borough name and total number of incidents.
numbers_by_boro <- as.data.frame(table(shooting_data$BORO)) 
names(numbers_by_boro)[c(1,2)] <- c("Boro", "Total")
# Plot the relationship
ggplot(data = numbers_by_boro, aes(Boro, Total)) +
       geom_point() + 
       labs(title = "NYC Shooting Incidents by Borough",
            x = "Borough",
            y = "Total incidents")
```

According to this graph, the most incidents occurred in Brooklyn and the Bronx, followed at a distance by Queens, Manhatten, and in the lowest set of incidents, Staten Island. 

Are Brooklyn and the Bronx more dangerous? Or are they simply far more populous? The original data did not include population figures for the five boroughs, so I will need to track this down to analyze further.

Per the data available from https://www.citypopulation.de/en/usa/newyorkcity/, as of the 2021 census, the boroughs have the following population:
```{r borough_population}
# Add population data from 2021 census
Pop <- c(1424948, 2641052, 1576876, 2331143, 493494)
# Include column for deaths relative to population
numbers_by_boro <- numbers_by_boro %>%
  mutate(Population = Pop, Shootings_Per_Thousand = Total*1000/Population)
# View new graph
ggplot(data = numbers_by_boro, aes(Boro, Shootings_Per_Thousand)) +
       geom_point() + 
       labs(title = "NYC Shooting Incidents by Borough",
            x = "Borough",
            y = "Shooting incidents per 1000 people")
```
With this new information, we can see that Bronx and Brooklyn do indeed have the most incidents by population but it is the Bronx rather than Brooklyn with the highest per-capita number of incidents. And due to its low population, Staten Island is pretty similar to Queens in terms of per-capita incidents.

## Further Questions
For future examination of this data set, here are the questions I have:
1. How do age, race, sex, education, and poverty play roles? 
- Do areas with higher levels of shooting have a high predominance of shootings by (and against) young males in poorer areas who may have less education? If so, it could suggest gang involvement.
- In contrast, in areas where victims tend to be wealthier and/or better educated, it may suggest a motive related to robbery.
- Clustered incidents where victims are disproportionately young (school age) suggests it may be a school shooting, whereas disparate cases of adult female victims may point more towards domestic violence issues.
2. How was the data collected? Why was there a gap in location description?
- As hypothesized earlier, the officers recording data may have considered location description unimportant but it certainly could not have been unknown in as many incidents as the N/A entry suggests. This is because those same incidents in general did NOT lack data on latitude and longitude, so clearly the location of the incident was known upon reporting.

## ADD BIAS IDENTIFICATION
TODO: Write conclusion, including any possible bias sources. Identify personal bias & how mitigating.
